---
title: "test tmap & mapview displaying interactive maps from earlier chunk"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

New issue 2021-05. On shinyapps learnr chunks display interactive maps from earlier chunks.

[Reproducible example code](https://github.com/afrimapr/afrilearnr/blob/main/inst/tutorials/test-tmap-cache/test-tmap-cache.Rmd)

Chunk1 makes an orange world map. 
Chunk2 makes a blue world map. 
Chunk3 makes an orange dot map. 
Chunk4 makes a blue dot map. 

If user Runs chunks 1-4 in that order, they all display map1.

If user runs chunk 3 then 4, both chunks display map3.

Happens on shinyapps but not locally.

Possibly related to issue fixed by RStudio in 2021-02 https://github.com/rstudio/learnr/issues/484
Causing us problems with our tutorials here : https://github.com/afrimapr/afrilearnr/issues/16


```{r tmap-test-orange, exercise = TRUE, exercise.eval = TRUE, warning=FALSE, message=FALSE}
#1
library(tmap)
tmap_mode("view")
data("World")

tm_shape(World) +
    tm_polygons("HPI")


```


```{r tmap-test-blue, exercise = TRUE, exercise.eval = TRUE, warning=FALSE, message=FALSE}
#2
library(tmap)
tmap_mode("view")
data("World")

tm_shape(World) +
    tm_polygons("HPI", palette='Blues')

```

### Testing similar in mapview

```{r mapview-test-orange, exercise = TRUE, exercise.eval = TRUE, warning=FALSE, message=FALSE}

#3
library(mapview)
mapview(breweries, col.regions='orange')


```


```{r mapview-test-blue, exercise = TRUE, exercise.eval = TRUE, warning=FALSE, message=FALSE}

#4
library(mapview)
mapview(breweries, col.regions='blue')

```

