```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r}
# Use install.packages("<package_name>") to install any of the packages below
library(tidyverse) ## For plotting and data wrangling.

library(leaflet) ## For leaflet interactive maps

library(sf) ## For spatial data

library(RColorBrewer) ## For colour palettes

library(htmltools) ## For html

library(leafsync) ## For placing plots side by side

library(kableExtra) ## Table output

```


# Birth Rates in North Carolina

* We will use the North Carolina (`nc`) data from the `sf` package.

* Let's load in the data for North Carolina using the function `st_read`.

```{r nc-data, eval = TRUE, messages = FALSE}

nc_df <- st_read(system.file("shape/nc.shp", package="sf"))

```

* `st` = spatial type and `.shp` is a common shape file format (e.g. GIS). 

## Rename the columns and view the data North Carolina

* Number of births for counties in North Carolina in 1974 

* Rename our columns to country, births, and geometry.


```{r nc-data2, eval = TRUE, messages = FALSE}

nc <- nc_df %>%
        select("NAME", "BIR74", "BIR79", "geometry") %>%
        rename("county" = "NAME", "births1974" = "BIR74", "births1979" = "BIR79")

```


## Let's inspect the data

* Let's load in the data for North Carolina (nc)

```{r nc-view, eval = TRUE}

head(nc)


```

## And inspect the structure


```{r nc-str, eval = TRUE}

str(nc)

```

## Building a map in ggplot2

```{r first-map1a, eval=TRUE}
ggplot(nc) #<<

```

```{r first-map1e, eval=TRUE}
ggplot(nc) +
  geom_sf(aes(fill = births1974)) +
  labs(title = "Births per county in 1974",
       x = "Longitude",
       y = "Latitude", 
       fill = "Births") +
  scale_y_continuous(breaks = 34:36) #<<
```

## Coordinate reference system

* Every location on earth is specified by a longitude and latitude. 

* The Coordinate Reference system (CRS) determines how the data will be projected onto a map. 

* We can check the CRS using `st_crs`:

```{r checking the CRS}

st_crs(nc)

```

* The CRS is specified in the attributes `epsg` and `proj4string`. 

## Transforming coordinate reference system


* You can transform a coordinate reference system using the `st_transform()`. 

* But what is a sensible coordinate reference system to assign? 

* Well, a good place to start is with the one that leaflet uses for plotting the world:  EPSG 4326.

```{r changing the CRS using st_transform}

nc <- st_transform(nc, "+init=epsg:4326")

st_crs(nc)

```

# Our first leaflet map

* Every plot starts with `leaflet()`

```{r first-leaflet-map1a, eval=TRUE}

leaflet(data = nc) 

```


* Layers are added using `%>%`

```{r first-leaflet-map1b, eval=TRUE}
leaflet(data = nc) %>%
  addTiles() #<<
```


* N.B. Layers are added with `%>%` in `leaflet` and `+` in `ggplot`. `%>%` also is used in the `tidyverse` packages. 


* We can set the view using `setView()`

```{r first-leaflet-map1c, eval=TRUE}
leaflet(data = nc) %>%
  addTiles() %>%
  setView(lng = 38, #<<
          lat = 0, #<<
          zoom = 2) #<<
```

* Add different background map using addProviderTiles

```{r first-leaflet-map1d, eval=TRUE}
leaflet(data = nc) %>%
  addProviderTiles(providers$Esri.OceanBasemap) %>%
  setView(lng = -80, 
          lat = 34.5, 
          zoom = 5)
```

* N.B. The different provider tiles come with different licensing.

* Add polygons using `addPolygons()`

```{r first-leaflet-map1e, eval=TRUE}
leaflet(data = nc) %>%
  addProviderTiles(providers$Stamen.Terrain) %>%
  setView(lng = -80, 
          lat = 34.5, 
          zoom = 5) %>%
  addPolygons() #<<
```

## Creating a colour palette

* What type of data are we showing? 

* Who is our audience? 

* Tools

  * [ColorBrewer2.org](https://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3)
  * [Color Picker](https://www.colorcodepicker.com/)


## Creating a colour palette

* `RColorBrewer` includes **sequential** colour palettes (e.g. number of people).

```{r pal-sequential, eval = TRUE}
display.brewer.all(type = "seq")
```

## Creating a colour palette

* `RColorBrewer` includes **diverging** colour palettes (e.g. to show distinct categories).

```{r pal-diverging, eval = TRUE}
display.brewer.all(type = "div")
```


* First we will define the colour palette and bins for the plot.

```{r Defining colour bins, eval = TRUE}

summary(nc$births1974)
summary(nc$births1979)

bins <- seq(0, 35000, 5000) 
```


* Then we can define the colours for the palette:


```{r Defining the palette, eval = TRUE}
pal74 <- colorBin("OrRd", domain = nc$births1974, bins = bins) 

pal79 <- colorBin("OrRd", domain = nc$births1979, bins = bins)
```


* Customising `addPolygons()`

```{r first-leaflet-map1f, eval=TRUE}
leaflet(data = nc) %>%
  addProviderTiles(providers$Stamen.Terrain) %>%
  setView(lng = -80, 
          lat = 34.5, 
          zoom = 6) %>%
  addPolygons(
    fillColor = ~pal74(nc$births1974),
    fillOpacity = 0.5, #<<
    color = "white", #<<
    opacity = 1, #<<
    weight = 2 #<<
  ) 
```

* Customising `addPolygons()`

```{r first-leaflet-map1g, eval=TRUE}
leaflet(data = nc) %>%
  addProviderTiles(providers$Stamen.Terrain) %>%
  setView(lng = -80, 
          lat = 34.5, 
          zoom = 6) %>%
  addPolygons(
    fillColor = ~pal74(nc$births1974), #<<
    fillOpacity = 1, #<<
    color = "blue", #<<
    opacity = 0.7, #<<
    weight = 1 #<<
  ) 
```

# What can you customise in addPolygons()

```{r what can you customise with addPolygons, eval = FALSE}

?addPolygons()

```

* `color:` stroke color
* `weight:` stroke width in pixels
* `opacity:` stroke opacity
* `fillColor:` fill color
* `fillOpacity:` fill opacity
* `highlightOptions:` Options for highlighting the shape on mouse over. 

* Let's assign our plot to an object.

```{r first-leaflet-map1h, eval=TRUE}
m1 <- leaflet(data = nc) %>% #<<
  addProviderTiles(providers$Stamen.Terrain) %>%
  setView(lng = -80, 
          lat = 34.5, 
          zoom = 6)

m1 %>%
  addPolygons(
    fillColor = ~pal74(nc$births1974), 
    fillOpacity = 0.7,
    opacity = 1,
    color = "white", 
    weight = 2) 


```

* Let's add some `highlightOptions`

```{r first-leaflet-map1j, eval=TRUE}
m1 %>%
  addPolygons(
      fillColor = ~pal74(nc$births1974), 
      fillOpacity = 0.7, 
      color = "white", 
      opacity = 1,
      weight = 2,
    highlight = highlightOptions( #<<
        weight = 3, #<<
        color = "blue", #<<
        fillOpacity = 1, #<<
        bringToFront = TRUE)) #<< 
```

## Let's add some labels!

`sprintf`: returns a character vector containing a formatted combination of text and variable values.


```{r Make our labels}
labels <- sprintf("<strong>%s</strong><br/>%g births", 
                  nc$county, nc$births1974) %>% lapply(htmltools::HTML)

head(labels, 1)
```

html - markup language for the web
* `<strong>` = bold; `<br/>` = new line

PHP - Hypertext Preprocessor
* `%s` = place holder for a character string; `%g` = general format place holder for a number

```{r first-leaflet-map1k, eval=TRUE}
(m1 <- m1 %>%
  addPolygons(data = nc,
      fillColor = ~pal74(nc$births1974),
      fillOpacity = 0.7,
      color = "white",
      opacity = 1,
      weight = 2,
      highlight = highlightOptions(
        weight = 3,
        color = "blue",
        fillOpacity = 1,
        bringToFront = TRUE),
      label = labels)) #<< 

```

* Let's add a legend

```{r first-leaflet-map1l, eval=TRUE}
m1 <- m1 %>%
  addLegend( #<<
    position = "bottomright", #<<
    pal = pal74, #<<
    values = ~nc$births1974, #<<
    title = "Births by county in 1974", #<<
    opacity = 1) #<<

m1
```

## Let's create a second map

* Let's create a second map of births in 1979.

* First we'll need to create a new set of labels

```{r labels for second map}

labels79 <- sprintf(
  "<strong>%s</strong><br/>%g births",
  nc$county, nc$births1979
) %>% lapply(htmltools::HTML)


```

## Let's create a second map

```{r second map}

m2 <- leaflet(data = nc) %>%
  addProviderTiles(providers$Stamen.Terrain) %>%
  setView(lng = -80, lat = 34.5, zoom = 6) %>%
  addPolygons(
      fillColor = ~pal79(nc$births1979),
      fillOpacity = 0.7,
      color = "white",
      opacity = 1,
      weight = 2,
      highlight = highlightOptions(
        weight = 3,
        color = "blue",
        fillOpacity = 1,
        bringToFront = TRUE),
      label = labels79)

```

```{r leaflet-map2a, eval=TRUE}
(m2 <- m2 %>%
  addLegend(
    position = "bottomright",
    pal = pal79,
    values = ~nc$births1979,
    title = "Births by country in 1979",
    opacity = 1))

```

## Placing two maps side by side

```{r leaflet-map3}

leafsync::sync(m1, m2, ncol = 2, sync = "all")
```

# Leaflet maps with points

```{r leaflet-map-points, echo = TRUE}

work <- data.frame(
  "location" = c("Valparaíso, Chile", 
                 "Curitiba, Brasil", 
                 "Sable Island, Nova Scotia", 
                 "Greifswald, Germany",
                 "Arusha, Tanzania",
                 "Kigali, Rwanda",
                 "Kingston, Jamaica",
                 "Asunción, Paraguay",
                 "East Kilbride, Scotland"), 
  "institute" = c("Instituto de Fomento Pesquero", 
                  "Universidade Federal do Paraná", 
                  "Dalhousie University", 
                  "Friedrich Loeffler Institut",
                  "Nelson Mandela African Institute of Science and Technology",
                  "National Institute of Statistics Rwanda",
                  "Caribbean National Statistical Offices",
                  "El Ministerio de Salud Pública y Bienestar Social y el Ministerio de Educación y Ciencias - Paraguay",
                  "Foreign Commonwealth Development Office"), 
  "work" = c("Chilean Pink Cusk Eel", 
             "Fox rabies", 
             "Grey seals", 
             "Fox rabies",
             "Teaching",
             "Teaching",
             "Teaching",
             "Teaching",
             "Teaching"), 
  "lat" = c(-33.0472, 
            -25.4290, 
            43.9337, 
            54.0865,
            -3.3995,
            -1.9415,
            18.0179,
            -25.2637,
            55.760869), 
  "lon" = c(-71.6127, 
            -49.2671, 
            -59.9149, 
            13.3923,
            36.7968,
            30.0574,
            -76.8099,
            -57.5759,
            -4.22407), 
  "icon" = c("fish", 
             "disease", 
             "gps", 
             "disease",
             "training",
             "training",
             "training",
             "training", 
             "training"))
```

```{r data-preview} 

work %>%
  kbl() %>%
  kable_paper("hover")

```


```{r work-leaflet, eval=TRUE}
leaflet(work) %>% 
      addProviderTiles(providers$Stamen.Watercolor) %>%
      addProviderTiles(providers$Stamen.TerrainLabels) %>%
      addCircleMarkers(~lon, ~lat)
```

* **Your Turn*** What else can you change about addCircleMarkers? Hint: Type `??addControl` Try adding: `clusterOptions = markerClusterOptions()` to your map. 

* Add labels

```{r point-labels}


labels <- sprintf(
  "<strong>%s</strong>",
  work$institute) %>% lapply(htmltools::HTML)

```

```{r}
(work_map <- leaflet(work) %>% 
                addProviderTiles(providers$Stamen.Watercolor) %>%
                addProviderTiles(providers$Stamen.TerrainLabels) %>%
                addCircleMarkers(~lon, ~lat, popup = ~labels))
```
