---
title: "How to get your own spatial data into R"
output: learnr::tutorial
runtime: shiny_prerendered
description: >
  An entry level outline to help you get your own data, of different types, into R as a first step to mapping it.
---

```{r setup, include=FALSE}
library(learnr)

# default options for all chunks
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE )

library(sf)               # working with vector data
library(afrilearndata)    # example data
#library(tmap)             # static and interactive mapping
library(raster)           # raster manipulation
#library(dplyr)           # data wrangling
library(mapview)          # interactive mapping
library(readr)            # reading text files

library(rgdal)            # to try to correct raster error on shinyapps

```

## A. Outline of this tutorial

This entry level tutorial will demonstrate getting spatial data of different types into R. The aim is to support you getting your own data into R before making maps or other plots.


### Outline of afrilearnr

This tutorial is part of the [afrilearnr](https://github.com/afrimapr/afrilearnr) package containing tutorials to teach spatial in R with African data. It is part of the [afrimapr](https://afrimapr.github.io/afrimapr.website/) project.

Through the magic of [learnr](https://rstudio.github.io/learnr/) you can modify the R code in the boxes below and press run to see results.

If you are accessing this from shinyapps you can also install [afrilearnr from github](https://github.com/afrimapr/afrilearnr) and run these tutorials locally.

A third option, for lower bandwidths, is to download a [pdf of the tutorial](https://github.com/afrimapr/afrilearnr/tree/main/inst/pdfs) and run through by copying the code into an R console.

### How this tutorial relates to others in afrilearnr

tutorial name        | outline                                   | recommended order
-------------------- | ----------------------------------------- | ------------------
`intro-to-spatial-r` |   an introduction to spatial data in R | 1
`get-my-data-in`     |    getting your own spatial data into R | 2 this one
`join-admin`         |    dealing with data referenced by names rather than coordinates | 3
`afrilearnr-crash-course` |   gallery of plots & code with minimal explanation | 4

### How to use this tutorial 

Click 'Next Topic' to move between sections, or selection section headings on the left. If you want to return the tutorial to its original state you can press 'Start Over' on the lower left.

To repeat these steps locally you would need the following packages.

```{r packages, echo=TRUE}

library(readr)            # reading text files
library(sf)               # working with vector data
library(mapview)          # interactive mapping
library(raster)           # raster manipulation
library(afrilearndata)    # example data, not required if using only your own data

```

## B. .csv, .txt or .xls file with coordinates

Text files containing point data are one of the commonest file types that we see in small-scale operational mapping of data. Usually these consist of one row per record (e.g. the location of a health facility or disease case or dwelling) with two columns containing the coordinates of the location (e.g. longitude & latitude or x & y), and other columns containing attributes of that location (e.g. facility or disease type). 

attribute1  | longitude | latitude
----------- | --------- | ------------------
location1   | -10 | 20
location2   | 10  | 0  
...         |     |

These files can be `.csv` comma delimited, or `.txt` space delimited or various spreadsheet formats including `.xls`.

To map these data in R usually requires a 3 step process.

1. read the data file into an R dataframe
1. convert the dataframe into an R spatial (package `sf`) object
1. plot the `sf` object

Here we will demonstrate the 3 steps using some airport data from the excellent [ourairports](https://ourairports.com/continents/AF/airports.csv) that we have extracted and saved in the [afrilearndata](https://github.com/afrimapr/afrilearndata) package. 

```{r csv-airports-setup}
# to avoid issues with interactive maps https://github.com/rstudio/learnr/issues/529
# IdSeed must be set to be unique in the Rmd, even for mode('plot') to allow user to change
#tmap_mode("plot")
htmlwidgets::setWidgetIdSeed(1)
```


```{r csv-airports, exercise = TRUE, exercise.eval = TRUE}

# 1. read into dataframe
filename <- system.file("extdata","afriairports.csv", package="afrilearndata", mustWork=TRUE)
mydf <- readr::read_csv(filename)

mydf <- mydf[(1:100), ] #select first 100 rows just to make quicker online

# 2. convert to sf object & set crs
mysf <- sf::st_as_sf(mydf, 
                     coords=c("longitude_deg", "latitude_deg"),
                     crs=4326)

# 3. quick interactive plot
mapview(mysf)    


```

  
  
**To apply the code chunk above to your own data :**

* set filename to the path to your file (this might just be something like `"mydata/myfile.csv"`) 
* replace `"longitude_deg", "latitude_deg"` with the names of the columns containing the coordinates in your data
* you may need to change `crs=4326` as explained below

### CRS
`crs` stands for Coordinate Reference System. It determines how coordinates are converted to a location on the Earth. 
In this case it tells `sf` what system to expect. In the majority of cases coordinates (e.g. collected from a GPS) are stored in a system represented by the code `4326`. 4326 is the EPSG code for longitude, latitude using the WGS84 datum, but you don't really need to know that. 4326 is a good number to remember !

**Question :** What happens when the `crs=4326` argument is not included in the code below ? Try adding it back in and see what the difference is.


```{r csv-crs-missing-setup}
# to avoid issues with interactive maps https://github.com/rstudio/learnr/issues/529
# IdSeed must be set to be unique in the Rmd, even for mode('plot') to allow user to change
#tmap_mode("plot")
htmlwidgets::setWidgetIdSeed(2)
```

```{r csv-crs-missing, exercise = TRUE, exercise.eval = TRUE, message = FALSE, warning = FALSE}

# 1. read into dataframe
filename <- system.file("extdata","afriairports.csv", package="afrilearndata", mustWork=TRUE)
mydf <- readr::read_csv(filename)

mydf <- mydf[(1:100), ] #select first 100 rows just to make quicker online

# 2. convert to sf object - NOTE crs missing
mysf <- sf::st_as_sf(mydf, 
                     coords=c("longitude_deg", "latitude_deg"))

# 3. quick interactive plot
mapview(mysf)    

```
  
  
**Answer :** You should see that when there is no `crs` argument the sf object is still created but mapview is unable to position it in the world. The points still appear but there is no map background.

### data structure

We can look at the structure of the original dataframe and the sf object using `names()` to show the column names and `head()` which returns the first six rows. What is the difference between them ? 

```{r dataframe-airports, echo=TRUE}

# original dataframe
names(mydf)
# sf object
names(mysf)  
# original dataframe
head(mydf)
# sf object
head(mysf)

```

  
You should see that the columns containing coordinates in the original dataframe are no longer there. In the new `sf` object there is a new column called `geometry` at the end that stores the spatial information. This demonstrates that an `sf` object behaves like a dataframe in many ways.  


### .xls files

For Microsoft Excel files you just need to change step 1 of the three step approach. You can read an excel file into a dataframe using the package [readxl](https://readxl.tidyverse.org/) with something like `readxl::read_excel(filename)`. Another option is to save the sheet that you want as a .csv file from MS Excel itself.


## C. Directly create an R object

An alternative is directly to create a dataframe within R containing coordinates. This is similar to the approach from the previous section except that dataframe creation replaces file reading at step 1.

In the example below try changing the coordinates within the dataframe at step 1, and run to see the points change.

```{r dataframe-sf-setup}
# to avoid issues with interactive maps https://github.com/rstudio/learnr/issues/529
# IdSeed must be set to be unique in the Rmd, even for mode('plot') to allow user to change
#tmap_mode("plot")
htmlwidgets::setWidgetIdSeed(3)
```

```{r dataframe-sf, exercise = TRUE, exercise.eval = TRUE, message = FALSE, warning = FALSE}

# 1. create dataframe
mydf <- data.frame(x=c(-10,10,30),
                   y=c(20,0,-20),
                   attribute=c("a","b","c"))

# 2. convert to sf object
mysf <- sf::st_as_sf(mydf, 
                     coords=c("x", "y"),
                     crs=4326)

# 3. quick interactive plot
mapview(mysf)    

```
  
  
Note that in this example the coordinates are stored in columns named x & y, which is passed to `sf::st_as_sf` as `coords=c("x", "y")`. To find out more about the arguments for any function you can type `?` and the function name e.g `?st_as_sf`



## D. Shapefiles (.shp)

Shapefiles continue to be a common format for spatial data despite the fact that they are rather old now and some things about them are not ideal. One thing that can confuse users is that a shapefile consists of a collection of files with the same name and different suffixes. If some of the files are not present then it may no longer be possible to get at the data.

e.g. myfile.shp, myfile.shx, myfile.dbf, myfile.prj

If a colleague emails use just a single file named `*.shp` then you will not be able to map it in R. You would need to ask them to email you all of the files.

Shapefiles can store points, lines or polygons. The example below uses a shapefile containing polygons.

```{r shp-countries-setup}
# to avoid issues with interactive maps https://github.com/rstudio/learnr/issues/529
# IdSeed must be set to be unique in the Rmd, even for mode('plot') to allow user to change
#tmap_mode("plot")
htmlwidgets::setWidgetIdSeed(4)
```

```{r shp-countries, exercise = TRUE, exercise.eval = TRUE, message = FALSE, warning = FALSE}

# read file into a spatial object
filename <- system.file("extdata","africountries.shp", package="afrilearndata", mustWork=TRUE)
africountries <- sf::read_sf(filename)

# quick interactive plot
mapview(africountries)

```
  
  
Because shapefiles are spatial files they can be read directly into a spatial (`sf`) object in R with `sf::read_sf(filename)`. This combines steps 1 & 2 from the csv example. In addition you don't need to specify in R which columns contain the coordinates or what the Coordinate Reference System (crs) is. This is effectively because these two steps will have been done when the file was created.


## E. .kml, .gpkg & .json

For other spatial vector formats (e.g. kml, geopackage & geojson) the same approach as for a shapefile usually works i.e. `sf::read_sf(filename)`.

Here we show an example with a .kml file of the simplified African highway network.

```{r kml-highways-setup}
# to avoid issues with interactive maps https://github.com/rstudio/learnr/issues/529
# IdSeed must be set to be unique in the Rmd, even for mode('plot') to allow user to change
#tmap_mode("plot")
htmlwidgets::setWidgetIdSeed(5)
```

```{r kml-highways, exercise = TRUE, exercise.eval = FALSE, message = FALSE, warning = FALSE}

filename <- system.file("extdata","trans-african-highway.kml", package="afrilearndata", mustWork=TRUE)

afrihighway <- sf::read_sf(filename)

# quick interactive plot
mapview(afrihighway)

```


## F. raster tiff 

To read in raster data we need to use the package `raster` instead of `sf`. The reading function in `raster` is also called `raster`. To read in a file use `myrast <- raster::raster(filename)` or just `myrast <- raster(filename)`. Similar to vector formats you can also use `mapview` to give a quick view of raster objects by simply passing the object name e.g. `mapview(myrast)`.

`raster(filename)` will also work with other raster formats such as ascii grids or .jpg.

```{r rast1-setup}
# to avoid issues with interactive maps https://github.com/rstudio/learnr/issues/529
# IdSeed must be set to be unique in the Rmd, even for mode('plot') to allow user to change
#tmap_mode("plot")
htmlwidgets::setWidgetIdSeed(6)
```

```{r rast1, exercise = TRUE, exercise.eval = TRUE, message = FALSE, warning = FALSE}

filename <- system.file("extdata","afripop2020.tif", package="afrilearndata", mustWork=TRUE)

myrast <- raster::raster(filename)

# quick interactive plot
mapview(myrast)

```
  
Note that the map above appears mostly dark. This is the same issue we came across in the intro-to-spatial-r tutorial. This is because there are few very high density cells and a majority of cells with very low values. This is a common issue with population data. The default, equal-interval, classification doesn't work well, most of the map falls in the lowest category. If you look very closely you can see a few very high value cells e.g. in Lagos & Cairo. 

In intro-to-spatial-r we fixed this in `tmap` using the `breaks=` argument to set the breakpoints between colours. In `mapview` we can achieve the same using `at=`.

**To try** replace the final line above with this : `mapview(myrast, at=c(0,1,10,100,1000,10000,100000))`. Experiment with different breakpoints. 


## G. `mapview` options

In these examples we have used `mapview` to give us a quick view by passing it only the spatial object and not specifying any other options. `mapview` is very flexible and by passing just a few arguments the map can be made much more informative. Try copy & pasting this line to replace the final line in the code window below and running it. It uses the columns named `type` and `name` from the datafile to colour and label the points. `cex` sets the size of the points, in this case making them smaller.
`mapview(mysf, zcol='type', label='name', cex=2)`

```{r mapview-options-airports-setup}
# to avoid issues with interactive maps https://github.com/rstudio/learnr/issues/529
# IdSeed must be set to be unique in the Rmd, even for mode('plot') to allow user to change
#tmap_mode("plot")
htmlwidgets::setWidgetIdSeed(7)
```

```{r mapview-options-airports, exercise = TRUE, exercise.eval = FALSE}

# 1. read into dataframe
filename <- system.file("extdata","afriairports.csv", package="afrilearndata", mustWork=TRUE)
mydf <- readr::read_csv(filename)

mydf <- mydf[(1:100), ] #select first 100 rows just to make quicker online

# 2. convert to sf object
mysf <- sf::st_as_sf(mydf, 
                     coords=c("longitude_deg", "latitude_deg"),
                     crs=4326)

# 3. quick interactive plot
mapview(mysf)    

```

To find out more about `mapview` options, type just `?mapview` into the window above and press run. This should display the manual page for the `mapview` function, scroll down to where you see **Arguments** in bold and it gives more information about settings. Note that not all of these are available for `sf` vector files.

These `mapview` arguments are the most useful :

argument  | value | what does it do ?
----------- | --------- | -------------------------------------
zcol   | a column name | determines how features are coloured and the legend
label   | a column name or some text  | gives a label that appears when mouse is hovered over
cex     | number e.g. 2 or a column name | sets point size to a constant number or the value held in a column 
col.regions | `'blue'` | a colour palette or individual colour for circle interiors
color | `'red'` | a colour palette or individual colour for circle borders
alpha.regions   | a number between 0 & 1 | opacity of the circle interior, 0=invisible 
alpha   | a number between 0 & 1 | opacity of the circle outline, 0=invisible which removes circle border and can be effective
legend | `TRUE` or `FALSE` | whether to plot a legend, TRUE by default
map.types | `c('CartoDB.Positron', 'OpenStreetMap.HOT')` | background map layer options
at | a series of numeric values e.g. `c(0,1,10)` | breakpoints between colours

Try modifying the airports maps above using some of these options.


## Next steps

More information on reading different spatial file formats into R can be found in [this section in the excellent Geocomputation in R](https://geocompr.robinlovelace.net/read-write.html#data-input)

## Summary

We hope you've enjoyed this brief intro to getting your own spatial data into R.

We've shown you :

i. how to make a map from a coordinates text file
i. the importance of CRS (Coordinate Reference System) to place data on a world map
i. how to read in other spatial data files for vector and raster data
i. options for making mapview maps more useful

This is a start, there are plenty of other options for making maps in R. We will cover other options in later tutorials, have a look at the [afrilearnr-crash-course](https://andysouth.shinyapps.io/afrilearnr-crash-course/) for some potential next steps. We welcome feedback, different ways of [getting in touch are on our website](https://afrimapr.github.io/afrimapr.website/get-involved/).

There is a short quiz if you'd like to test your learning. 


## Quiz

```{r quiz}
quiz(

  question("What is the order of steps when reading in a text file of coordinates ?",
    answer("1.read into dataframe 2.convert to sf object & set crs 3.quick interactive plot", correct = TRUE),
    answer("1.convert to sf object & set crs 1.read into dataframe 3.quick interactive plot"),
    answer("1.read into dataframe 2.quick interactive plot 3.convert to sf object & set crs"),
    answer("1.quick interactive plot 2.read into dataframe 3.convert to sf object & set crs"),
    allow_retry = TRUE
  ),   
  question("What is the function to convert a dataframe to spatial sf format ?",
    answer("sf::", message = "this is an optional indication that a function is from the `sf` package, it can help to make code more readable, it will not do anything on its own"),
    answer("st_as_sf()", correct = TRUE),
    answer("st_read()", message = "`st_read` is used to read in files that are already in a spatial format e.g. `.shp`"),
    answer("read_csv()", message = "this reads in a text file but does not convert it to spatial, sf, format")
  ),
  question("What is the most common crs (coordinate reference system) for longitude, latitude data  ?",
    answer("c('longitude', 'latitude')", message = "this specifies the columns containing coordinates, not the crs"),
    answer("4326", correct = TRUE),
    answer("'4326'", message = "this is the correct number, but when placed in quotes R will intrepret it as text and it won't work as expected"),   
    # adding the answer below didn't do what I wanted - it only gave correct if both options were selected - I find it a bit starge that wgs84 works anyway, so best to leave out
    # answer("wgs84", correct = TRUE, message="this works too, but we recommend you use 4326"), 
    answer("EPSG", message = "EPSG is the name of the coding ssytem, not the code itself"),
    allow_retry = TRUE
  ),    
  question("In a coordinates file what names can be used for the columns containing coordinates ?",
    answer("longitude,latitude", correct = TRUE),
    answer("x,y", correct = TRUE),    
    answer("long,lat", correct = TRUE),
    answer("horizontal,vertical", correct = TRUE),
    answer("salt,pepper", correct = TRUE),
    allow_retry = TRUE,
    incorrect = "Any names can be used for columns as long as they are specified in st_as_sf(coords=c('x','y'). However, it is a good idea to stick to conventional names such as longitude,latitude to avoid confusion"
  )
)
```

```{r echo=FALSE}
# chunk below enables printing whole tutorial from browser e.g. to pdf
# DO NOT put any #comments in the chunk below, that stops it from working !! 
# from https://github.com/rstudio/learnr/issues/465
# saving csss in a separate file print.css didn't work locally or on shinyapps because browser couldn't find file 
```
```{css echo=FALSE}
@media print {
  .topicsContainer,
  .topicActions,
  .exerciseActions .skip {
    display: none;
  }
  .topics .tutorialTitle,
  .topics .section.level2,
  .topics .section.level3:not(.hide) {
    display: block;
  }
  .topics {
    width: 100%;
  }
  .tutorial-exercise, .tutorial-question {
    page-break-inside: avoid;
  }
  .section.level3.done h3 {
    padding-left: 0;
    background-image: none;
  }
  .topics .showSkip .exerciseActions::before {
    content: "Topic not yet completed...";
    font-style: italic;
  }
}  
  
```
