---
title: "Blank afrimapr tutorial template"
output: learnr::tutorial
runtime: shiny_prerendered
description: >
  A blank template for afrimapr authors to create other tutorials in a consistent format
---

```{r setup, include=FALSE}
library(learnr)

# default options for all chunks
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE )

#[TODO packages in here not shown to the user]

library(sf)               # working with vector data
library(afrilearndata)    # example data
library(tmap)             # static and interactive mapping
library(raster)           # raster manipulation
#library(dplyr)           # data wrangling
#library(mapview)         # interactive mapping

library(rgdal)            # to try to correct raster error on shinyapps

```


## Outline of this tutorial

[TODO: ADD A FEW SENTENCES INTRO TO THIS TUTORIAL]


### Outline of afrilearnr

This tutorial is part of the [afrilearnr](https://github.com/afrimapr/afrilearnr) package containing tutorials to teach spatial in R with African data. It is part of the [afrimapr](https://afrimapr.github.io/afrimapr.website/) project.

Through the magic of [learnr](https://rstudio.github.io/learnr/) you can modify the R code in the boxes below and press run to see results.

If you are accessing this from shinyapps you can also install [afrilearnr from github](https://github.com/afrimapr/afrilearnr) and run these tutorials locally.

A third option, for lower bandwidths, is to download a [pdf of the tutorial](https://github.com/afrimapr/afrilearnr/tree/main/inst/pdfs) and run through by copying the code into an R console.

### How this tutorial relates to others in afrilearnr

tutorial name        | outline                                   | recommended order
-------------------- | ----------------------------------------- | ------------------
`intro-to-spatial-r` |   an introduction to spatial data in R | 1 this one
`get-my-data-in`     |    getting your own spatial data into R | 2
`join-admin`         |    dealing with data referenced by names rather than coordinates | 3
`afrilearnr-crash-course` |   gallery of plots & code with minimal explanation | 4

### How to use this tutorial 

[TODO add a few sentences about using this tutorial]

To repeat these steps locally you would need the following packages.

[TODO packages in here will be shown to the user]
```{r packages, echo=TRUE}

library(readr)            # reading text files
library(sf)               # working with vector data
library(mapview)          # interactive mapping
library(raster)           # raster manipulation
library(afrilearndata)    # example data, not required if using only your own data

```

## [TODO create sections using ##]

```{r template-exercise, exercise = TRUE, exercise.eval = TRUE}

# TODO add exercise code here    

```

## Other sources of information

[Geocomputation in R](https://geocompr.robinlovelace.net/)



```{r echo=FALSE}
# chunk below enables printing whole tutorial from browser e.g. to pdf
# DO NOT put any #comments in the chunk below, that stops it from working !! 
# from https://github.com/rstudio/learnr/issues/465
# saving csss in a separate file print.css didn't work locally or on shinyapps because browser couldn't find file 
```
```{css echo=FALSE}
@media print {
  .topicsContainer,
  .topicActions,
  .exerciseActions .skip {
    display: none;
  }
  .topics .tutorialTitle,
  .topics .section.level2,
  .topics .section.level3:not(.hide) {
    display: block;
  }
  .topics {
    width: 100%;
  }
  .tutorial-exercise, .tutorial-question {
    page-break-inside: avoid;
  }
  .section.level3.done h3 {
    padding-left: 0;
    background-image: none;
  }
  .topics .showSkip .exerciseActions::before {
    content: "Topic not yet completed...";
    font-style: italic;
  }
}  
  
```
