---
title: "test tmap failing in view mode when run by learner"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

This chunk - the Hello World! example from tmap, with tmap_mode("view") added, works when the tutorial is loaded, or when run from console, but used to fail (with $ operator is invalid for atomic vectors) when 'Run Code' is pressed. This was fixed by RStudio in 2021-02 https://github.com/rstudio/learnr/issues/484


```{r tmap-test, exercise = TRUE, exercise.eval = TRUE, message = FALSE, warning = FALSE}

library(tmap)

data("World")

tmap_mode("view")

tm_shape(World) +
    tm_polygons("HPI")


```

Similar chunk loses newlines after `tmap_mode("view")` and initially fails in a different way on 'Run Code', but when the newlines are added back in fails same as above.  
```{r tmap-test2, exercise = TRUE, exercise.eval = TRUE}

library(tmap)

data("World")

tmap_mode("view")

tm <- tm_shape(World) +
    tm_polygons("HPI")

ptm <- print(tm)

#suggestion Martijn
tmap_leaflet(tm)

```
